ASAH1 rGBM Manuscript ELISA Figure

### 2023-08-26 results/elisa_plasma_proteomics/manuscript_elisa_figure_analysis.Rmd
exploration_assoc_mindy_saav_w_ttr

<!-- # Data File Paths -->
```{r message=FALSE, warning=FALSE, include=FALSE}
patient_clinical_data <- "../../data/elisa_plasma_proteomics/Patients_information_FINAL.csv"
patient_longitudinal_mapping_file <- "../../data/elisa_plasma_proteomics/Patients_for_ELISA_longitudinal_mapping.csv"
elisa_data_file <- "../../data/elisa_plasma_proteomics/ELISA_data.csv"
```

@@

  gene_stat_frame <- parallel::mclapply(split(data_sub, data_sub$gene), function(.d){
    if (method=="paired.perm.test"){
      .d %>%
        dplyr::select(-gene) %>%
        tidyr::pivot_wider(id_cols = "Number.x", names_from = "Manifestation", values_from = "log2_conc") -> .d_wide
    } else {
      .d -> .d_wide
    }
    message(sprintf("INFO: Computing %s for %s for comparison %s", method, unique(.d$gene), paste(comp, collapse=", ") ))
    do_perm_test(data = .d_wide, gene = .d$gene, response_var = "log2_conc", group1 = comp[1], group2 = comp[2], method = method, grouping_column = "Manifestation")
  }, mc.cores = 4) %>%
    rbindlist() %>%
    unique()
}, mc.cores=3) %>% rbindlist()

@@

  gene_stat_frame <- parallel::mclapply(split(data_sub, data_sub$gene), function(.d){
    if (method=="paired.perm.test"){
      .d %>%
        dplyr::select(-gene) %>%
        tidyr::pivot_wider(id_cols = "Number.x", names_from = "collection_type2", values_from = "log2_conc") -> .d_wide
    } else {
      .d -> .d_wide
    }
    message(sprintf("INFO: Computing %s for %s for comparison %s", method, unique(.d$gene), paste(comp, collapse=", ") ))
    do_perm_test(data = .d_wide, gene = .d$gene, response_var = "log2_conc", group1 = comp[1], group2 = comp[2], method = method, grouping_column = "collection_type2")
  }, mc.cores = 4) %>%
    rbindlist() %>%
    unique()
}, mc.cores=3) %>% rbindlist()

@@

## ROC Figure 
g <- ggplot(pred_obj_3$roc_data, aes(m=GBM, d=factor(obs, levels = c("ctrl", "GBM")), color=Model)) + 
  geom_roc(n.cuts=0) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_3$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
  coord_equal() +
  style_roc(x="1 - Specificity", y="Sensitivity")

@@

<!-- # Prediction for iGBM vs rGBM ratio -->
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
# iGBM vs rGBM ratio ----
data_long %>% 
  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
  dplyr::group_by(gene, Number.x) %>%
  dplyr::group_map(function(.d, ...){
    .d <- as.data.table(.d)
    .d %>% 
      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>% 
      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
    # print(.d)
    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
      .d_initial <- .d[manifestation_type=="Initial manifestation"]
      .d_second <- .d[manifestation_type=="Second manifestation"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else if (nrow(.d)==2) {
      .d_initial <- .d[collection_type=="E1"]
      .d_second <- .d[collection_type=="E2"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else {
      tmp$ratio <- .d$conc
    }
    return(tmp)
  }, .keep = TRUE) %>%
  rbindlist() %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
  dplyr::mutate(log2_conc = log2(ratio)) %>%
  # dplyr::group_by(gene, Number.x) %>%
  # dplyr::add_count() %>%
  # dplyr::filter(n==2) %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata

@@

<!-- # Latency Data -->
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
latency_data_file <- "../../data/elisa_plasma_proteomics/ELISA_LATENCY_KM.csv"
latency_data <- fread(latency_data_file)
# Clean names
latency_data <- rstatix::make_clean_names(latency_data)
latency_data %>%
  dplyr::mutate(ASAH1.ng.ml = log2(ASAH1.ng.ml+1),
                SYNM.ng.ml = log2(SYNM.ng.ml+1),
                GPNMB.ng.ml = log2(GPNMB.ng.ml+1),
                MMP.9.ng.ml = log2(MMP.9.ng.ml+1)) %>%
  dplyr::mutate(ASAH1 = ifelse(ASAH1.ng.ml>=mean(ASAH1.ng.ml), "high", "low"),
                SYNM = ifelse(SYNM.ng.ml>=mean(SYNM.ng.ml), "high", "low"),
                GPNMB = ifelse(GPNMB.ng.ml>=mean(GPNMB.ng.ml), "high", "low"),
                MMP9 = ifelse(MMP.9.ng.ml>=mean(MMP.9.ng.ml, na.rm=T), "high", "low")) -> latency_data

@@

@@

## COXPH ----
fit <- coxph(Surv(Time.to.reccurence..days., status) ~ sex + ASAH1.ng.ml + SYNM.ng.ml + GPNMB.ng.ml + MMP.9.ng.ml, data = ttr_surv_data)
print(summary(fit))

>>>>

<!-- # Data File Paths -->
```{r message=FALSE, warning=FALSE, include=FALSE}
patient_clinical_data <- here::here("data/elisa_plasma_proteomics/Patients_information_FINAL.csv")

patient_longitudinal_mapping_file <- here::here("data/elisa_plasma_proteomics/Patients_for_ELISA_longitudinal_mapping.csv")

elisa_data_file <- here::here("data/elisa_plasma_proteomics/ELISA_data.csv")
```

@@

  gene_stat_frame <- parallel::mclapply(split(data_sub, data_sub$gene), function(.d){
    if (method=="paired.perm.test"){
      .d %>%
        dplyr::select(-gene) %>%
        tidyr::pivot_wider(id_cols = "Number.x", names_from = "Manifestation", values_from = "log2_conc") -> .d_wide
    } else {
      .d -> .d_wide
    }
    message(sprintf("INFO: Computing %s for %s for comparison %s", method, unique(.d$gene), paste(comp, collapse=", ") ))
    do_perm_test(data = .d_wide, gene = .d$gene, response_var = "log2_conc", group1 = comp[1], group2 = comp[2], method = method, grouping_column = "Manifestation")
  }) %>%
    rbindlist() %>%
    unique()
}) %>% rbindlist()

@@

  gene_stat_frame <- parallel::mclapply(split(data_sub, data_sub$gene), function(.d){
    if (method=="paired.perm.test"){
      .d %>%
        dplyr::select(-gene) %>%
        tidyr::pivot_wider(id_cols = "Number.x", names_from = "collection_type2", values_from = "log2_conc") -> .d_wide
    } else {
      .d -> .d_wide
    }
    message(sprintf("INFO: Computing %s for %s for comparison %s", method, unique(.d$gene), paste(comp, collapse=", ") ))
    do_perm_test(data = .d_wide, gene = .d$gene, response_var = "log2_conc", group1 = comp[1], group2 = comp[2], method = method, grouping_column = "collection_type2")
  }) %>%
    rbindlist() %>%
    unique()
}) %>% rbindlist()

@@

## ROC Figure 
g <- ggplot(pred_obj_3$roc_data, aes(m=GBM, d=factor(obs, levels = c("ctrl", "GBM")), color=Model)) + 
  geom_roc(n.cuts=0) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_3$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
  coord_equal() +
  style_roc(x = "1 - Specificity", 
            y = "Sensitivity")

@@

<!-- # Prediction for iGBM vs rGBM ratio -->
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
# iGBM vs rGBM ratio ----
data_long %>% 
  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
  dplyr::group_by(gene, Number.x) %>%
  dplyr::group_map(function(.d, ...){
    .d <- as.data.table(.d)
    .d %>% 
      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>% 
      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
    # print(.d)
    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
      .d_initial <- .d[manifestation_type=="Initial manifestation"]
      .d_second <- .d[manifestation_type=="Second manifestation"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else if (nrow(.d)==2) {
      .d_initial <- .d[collection_type=="E1"]
      .d_second <- .d[collection_type=="E2"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else {
      tmp$ratio <- .d$conc
    }
    return(tmp)
  }, .keep = TRUE) %>%
  rbindlist() %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
  dplyr::mutate(log2_conc = log2(ratio)) %>%
  # dplyr::group_by(gene, Number.x) %>%
  # dplyr::add_count() %>%
  # dplyr::filter(n==2) %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to = c("iGBMoverrGBM"))) -> pdata

@@

<!-- # Latency Data -->
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
latency_data_file <- here::here("data/elisa_plasma_proteomics/ELISA_LATENCY_KM.csv")
latency_data <- fread(latency_data_file)
# Clean names
latency_data <- rstatix::make_clean_names(latency_data)
latency_data %>%
  dplyr::mutate(ASAH1.ng.ml = log2(ASAH1.ng.ml+1),
                SYNM.ng.ml = log2(SYNM.ng.ml+1),
                GPNMB.ng.ml = log2(GPNMB.ng.ml+1),
                MMP.9.ng.ml = log2(MMP.9.ng.ml+1)) %>%
  dplyr::mutate(ASAH1 = ifelse(ASAH1.ng.ml>=mean(ASAH1.ng.ml), "high", "low"),
                SYNM = ifelse(SYNM.ng.ml>=mean(SYNM.ng.ml), "high", "low"),
                GPNMB = ifelse(GPNMB.ng.ml>=mean(GPNMB.ng.ml), "high", "low"),
                MMP9 = ifelse(MMP.9.ng.ml>=mean(MMP.9.ng.ml, na.rm=T), "high", "low")) -> latency_data

@@

library(readr)
write_tsv(ttr_surv_data,
          here::here("data/elisa_plasma_proteomics/ttr_survival_data.tsv"))

@@

## COXPH ----
ttr_surv_data
fit <- coxph(Surv(Time.to.reccurence..days., status) ~ sex + ASAH1.ng.ml + SYNM.ng.ml + GPNMB.ng.ml + MMP.9.ng.ml, data = ttr_surv_data)
print(summary(fit))

# Cox-PHM

This section was created as part of the revision process, to check for the effect of age and sex in the model.

```{r}
?Surv
?coxphmulti
```

```{r}
## COXPH ----
#ttr_surv_data
fit2 <- coxph(Surv(Time.to.reccurence..days., status) ~ sex + Age.at.surgery..years. + ASAH1.ng.ml + SYNM.ng.ml + GPNMB.ng.ml + MMP.9.ng.ml, data = ttr_surv_data)
print(summary(fit2))
